---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# RTransferEntropy

The goal of RTransferEntropy is to implement transfer entropy for the Shannon and the Renyi-methodology.

## Installation

You can install RTransferEntropy from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("BZPaper/Transfer-Entropy")
```

## Example using simulated data

Simulate a simple model to obtain two time series that are not independent (see simulation study in Dimpfl and Peter (2013)),
i.e. one time series is lag of the other plus noise. In this case, one expects significant information flow from x to y 
and none from y to x.

### Simulating a Time-Series
```{r}
library(RTransferEntropy)
set.seed(20180108)
n <- 100000
x <- rep(0, n + 1)
y <- rep(0, n + 1)

for (i in seq(n)) {
  x[i + 1] <- 0.2 * x[i] + rnorm(1, 0, 2)
  y[i + 1] <- x[i] + rnorm(1, 0, 2)
}

x <- x[-1]
y <- y[-1]
```

### Visualisation
```{r}
library(data.table)
library(ggplot2)
library(latex2exp)
theme_set(theme_light())

df <- rbindlist(list(
  data.table(x, y, grp = "Contemporaneous Effect"),
  data.table(shift(x, 1), y, grp = "Time-Delayed Effect")
))

ggplot(df, aes(x = x, y = y, color = grp)) + 
  geom_point(alpha = 0.1) +
  geom_smooth() +
  facet_wrap(~grp) +
  labs(x = TeX("Left: x_t Right: x_{t-1}"), y = TeX("y_t"), 
       title = "Scatterplot of the Time-Series") +
  scale_color_brewer(palette = "Set1", guide = F)
```

### Shuffled Transfer Entropy
```{r}
x_code <- code_sample(x)
y_code <- code_sample(y)

set.seed(20180108 + 1)
(shuffled_TE <- shuffled_transfer_entropy(x = x_code, lx = 1,
                                          y = y_code, ly = 1))

```

### Shannon Transfer Entropy
```{r}
set.seed(20180108 + 1)
(Shannon_TE <- transfer_entropy(x = x_code,
                                lx = 1,
                                y = y_code,
                                ly = 1,
                                q = 0.5,
                                entropy = "Shannon",
                                shuffle = TRUE,
                                const = FALSE,
                                constx = 0,
                                consty = 0,
                                nreps = 2,
                                shuffles = 6,
                                ncores = parallel::detectCores() - 1,
                                type = "quantiles",
                                quantiles = c(5, 95),
                                bins = NULL,
                                limits = NULL,
                                nboot = 10,
                                burn = 50))
```
