---
title: "Vignette Title"
author: "Vignette Author"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style
Some Examples:

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures
We define the `mean_`-function as  (note the `eval=F` in the code: This code will not be run (the function in this case will not be sourced)! It will be displayed only!)
```{r part1, eval=F}
mean_ <- function(x) sum(x) / length(x)
```

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))



# Some David Stuff


Some Examples:

We define the `mean_`-function as  (note the `eval=F` in the code: This code will not be run (the function in this case will not be sourced)! It will be displayed only!)
```{r part1, eval=F}
mean_ <- function(x) sum(x) / length(x)
```

If we want to output something and show the code, we can leave the options empty

```{r part2}
x <- rnorm(10)
summary(x)
```

If we only want to show some output but do not want to show the code, we can use the `echo=F` option

```{r part3, echo=F}
plot(1:1000, rnorm(1000, sd = 0.01), type = "l", xlab = "Time", ylab = "Return")
```


# Equations

To write an equation:

\begin{equation}
  x = \sqrt{y + 1}
\end{equation}

or like this 

$$
  x = \sqrt{y + 1}
$$

